---
title             : "Review AI - Mental Healthcare Products"
shorttitle        : "Review AI Mental Health"

author: 
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "Anne-Kathrin.Kleine@psy.lmu.de"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : ""
    affiliation   : ""
    role:
      - ""
      - ""

affiliation:
  - id            : "1"
    institution   : "LMU"
  - id            : ""
    institution   : ""

authornote: |
  LMU

abstract: |
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "AI, Mental Health, Products"
wordcount         : "X"

bibliography      : "../config/LMU_AI_Team.bib"

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : 
  #papaja::apa6_pdf: default
  word_document: default 
header-includes:
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}
    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
csl               : "../config/apa.csl" #< path to csl file
documentclass     : "apa7"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, output_format = "all", encoding = encoding, output_dir = "../docs", output_file = "imp_issues") })
---

```{r setup, include = FALSE}
library("papaja")
library(googledrive)
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

## Introduction 

Although mental healthcare has been slower to adapt AI technology compared to physical healthcare [@miller_brown18, @jiang_etal17], the number of AI-powered mental health applications is on the rise [@vaidyam_etal19]. Two types of applications, developed for different user groups, can be distinguished. The first group consists of applications used by the general public or patient groups, such as chatbots, digital phenotyping tools, and artificially intelligent virtual agents [@vaidyam_etal19, @fiske_etal19, @lydon-staley_etal19]. The second group of applications is used by mental healthcare practitioners and aims at increasing the quality of therapeutic interventions [@bickman20]. \\

Although there have been claims to recognize the benefits of integrating AI into mental healthcare with the aim of enhancing diagnosis, treatment, and clinical administration accuracy and quality [@shatte_etal19], many tools that have been tested in research have never reached the production phase [@lee_etal21, @sendak_etal20, @chekroud_etal21]. 
This applies in particular to diagnostic and recommendation tools aiming at detecting psychological disorders and suggesting clinical treatment approaches (e.g., psychopharmacotherapy versus psychotherapy) [@chekroud_etal21]. 
Reasons for the lack of implementation into clinical practice include patient data confidentiality issues, a lack of understanding and resulting scepticism to use AI tools on the part of mental health practitioners and related explainablity versus performance trade-offs, and the frequency of reported erroneous predictions (e.g., among underrepresented groups) [[@roth_etal21, @chen_etal22, @sendak_etal20, @kelly_etal19, @chekroud_etal21, @aafjes-vandoorn_etal21]. 
In addition, mental disorders often have overlapping symptoms with other mental or physical conditions, adding an additional challenge to clear-cut diagnostic. \\

The current systematic review seeks to uncover the gap between the current state of research on the implementation of AI in mental health and the products available on the market.
To this end, we conduct a systematic review of reviews on 






\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
