---
title             : "Internet-based cognitive-behavioral therapy (ICBT): Approaches and difficulties"
shorttitle        : "ICBT - Approaches and difficulties"

author: 
  - name          : "Anne-Kathrin Kleine"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "Anne-Kathrin.Kleine@psy.lmu.de"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
  - name          : ""
    affiliation   : ""
    role:
      - ""
      - ""

affiliation:
  - id            : "1"
    institution   : "LMU"
  - id            : ""
    institution   : ""

authornote: |
  LMU

abstract: |
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "ICBT"
wordcount         : "X"

bibliography      : "../config/LMU_AI_Team.bib"

floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : 
  #papaja::apa6_pdf: default
  word_document: default 
header-includes:
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}
    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother
csl               : "../config/apa.csl" #< path to csl file
documentclass     : "apa7"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, output_format = "all", encoding = encoding, output_dir = "../docs", output_file = "imp_issues") })
---

```{r setup, include = FALSE}
library("papaja")
library(googledrive)
r_refs("r-references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

## ICBT: Effectiveness

- several studies prove effectiveness of different ICBT approaches [@andersson_etal19]
- differentiation between clinician-guided (ICBT-CG) and self-guided (ICBT-SG) [@dear_etal15]
- almost 100 ICBT apps [@martinengo_etal21]
- effectiveness of ICBT-SG also demonstrated in Germany [@kramer_etal22]
- meta-analysis ICBT-SG: similar deteriration rates as other CBT intervention; lower deterioration than control conditions [@karyotaki_etal18]
- however, some scepticism against ICBT in general (and ICBT-SG specifically?) [@ghaneirad_etal21]: 1/3 of participants did not participate in online therapy
    - main reasons: refused online intervention; technical difficulties 
    - Aus @ghaneirad_etal21: "In einer aktuellen Studie konnten Topooco et al. (2017) zeigen, dass geringes Wissen über digital gestützte Behandlungen aufseiten der Patienten und Behandler zur geringen Akzeptanz des Behandlungssettings führte. Auch äußerten die Behandler Besorgnis über die technische Durchführbarkeit und Sicherheit solcher Behandlungen (Waller et al. 2020). Um die Erfahrung und Akzeptanz dieses Behandlungssettings zu steigern, könnten die behandelnden Psychotherapeuten auch in regulären Therapien außerhalb von Lockdownzeiten zwischendurch videogestützte Therapieeinheiten einführen. Mit entsprechenden motivierenden psychotherapeutischen Interventionen könnte dies die Gruppe der verweigernden Patienten künftig erheblich reduzieren."
    
### Ideas: 
- evaluating use of ICBT-SG in Germany - who are the main users? Who refuses to use ICBT-SG? Why? 
- possibly collaboration with ICBT-SG provider; e.g., Charitee Berlin (see https://www.selfapy.com/ueber-uns/wissenschaft): what are the reasons for refusing online psychotherapy? How to reach particularly vulnerable sceptical groups (e.g. older less educated)? 

\newpage

# References

::: {#refs custom-style="Bibliography"}
:::
